Structural Variability

### 2016-04-12 INFORMATION/paper.tex
cambios 12/04/2016

@@

{\bf Selection of the reference protein:}
For each family of proteins we selected a reference ``wilt type'' protein, which we consider is the most structurally representative protein of the family. To get this protein, we first calculated the average structure of each multiple alignment. Then, we calculated the ``Mean Square Deviation'' (MSD) between the structure of each protein of the family and the obtained average structure. Finally, we selected as the reference protein the one with the lower value of MSD.

{\bf Alignments analysis:}
For each family of proteins, the reference protein was aligned with each of the other proteins of the family. Then, the aligned and nonaligned sites, their coordinates, and the sequence identity for each pair of proteins were obtained.

{\bf ENM of the reference protein:}

We considered the backbone fluctuations of the reference protein around its equilibrium conformation to be described by a coarse - grained ``Elastic Network Model'' (ENM), which represents a protein as a network of nodes placed at the alpha carbons ($C_{\alpha}$) connected by springs. In general, the ENM potential is of the form:

\begin{equation}
V(r) = 1/2*(r-r_{0})^{T}K(r-r_{0})
\end{equation}

where, for a protein of $N$ sites, $r$ is a column vector with $3N$ elements: the $x$, $y$, $z$ coordinates of the $N$ $C_{\alpha}$, $r_{0}$ is the equilibrium structure, and $K$ is the ``stiffness'' matrix, which represents the network\textsc{\char13} s topology and spring force constants. 
Specifically, we used the ``Anisotropic Network Model'' (ANM). Following this model, we gave a spring force constant of 1 to sites at a distance $\leq$ 10 \AA  and a spring force constant of 0 to sites at a distance > 10 \AA. We selected 10 \AA  as the cut off value after optimization using a range of values form 8 to 12 (data not shown). This cut off value is also the most commonly used.

\section{Conclusion}
Not much of a paper, but it's a start.

>>>>

{\bf ENM of a reference protein:}

We considered the backbone fluctuations of the reference ``wilt type'' protein around its equilibrium conformation to be described by a coarse - grained ``Elastic Network Model'' (ENM), which represents a protein as a network of nodes placed at the alpha carbons ($C_{\alpha}$) connected by springs. In general, the ENM potential is of the form:

\begin{equation}
V_{wt}(\mathbf{r}) = 1/2(\mathbf{r}-\mathbf{r}^{0}_{wt})^{T}\mathbf{K}_{wt}(\mathbf{r}-\mathbf{r}^{0}_{wt})
\end{equation}

where, for a protein of $N$ sites, $\mathbf{r}$ is a column vector with $3N$ elements: the $x$, $y$, $z$ coordinates of the $N$ $C_{\alpha}$, $\mathbf{r}^{0}_{wt}$ is the equilibrium structure of the reference protein, and $\mathbf{K}_{wt}$ is the ``stiffness'' matrix, which represents the network\textsc{\char13} s topology and spring force constants.

{\bf LF-ENM model of mutant proteins:}

To simulate mutants of each reference protein and thus to generate datasets of theoretical proteins, we used the ``Linearly Forced Elastic Network Model'' (LF - ENM). This model predicts the effect of a single mutation adding a linear perturbative term to the reference potential:

\begin{equation}
V_{mut}=V_{wt} - \mathbf{f}^{T}(\mathbf{r}-\mathbf{r}^{0}_{wt})
\end{equation}

where $\mathbf{f}$ is a column vector with $3N$ elements that models the mutation. The equilibrium structure of the mutant $\mathbf{r}^{0}_{mut}$ is the value of $\mathbf{r}$ that minimizes $V_{mut}$. Using Eqs. (1) and (2) we find the structural variation due to the mutation:

\begin{equation}
d\mathbf{r}^{0} \equiv  \mathbf{r}^{0}_{mut} - \mathbf{r}^{0}_{wt} = \mathbf{K}^{-1}_{wt}\mathbf{f}
\end{equation}

This equation shows that the structural perturbation introduced by a mutation is related to the mutation and to the network of oscillators, via $\mathbf{K}^{-1}_{wt}$. We should note here that $\mathbf{K}^{-1}_{wt}$ is actually the pseudo inverse, because $\mathbf{K}_{wt}$ has six zero eigenvalues, corresponding to translations and rotations, so that it is not invertible.

{\bf LF-ENM for multiple mutants: }
To simulate proteins with not one but multiple mutations we considered additive mutations by the assumption that $\mathbf{K}^{mut}$ \cong $\mathbf{K}^{wt}$:

\begin{equation}
d\mathbf{r}^{tot} = d\mathbf{r}^{mut1}_{wt}+d\mathbf{r}^{mut2}_{wt}+ ...+d\mathbf{r}^{mutm}_{wt}+...+d\mathbf{r}^{mutNmut}_{wt}
\end{equation}

\begin{equation}
d\mathbf{r}^{tot} = =\mathbf{K}^{-1}_{wt}\times(\mathbf{fmut1}+\mathbf{fmut2}+...+\mathbf{fmutm}+...+\mathbf{fmutNmut}) 
\end{equation}

Where $d\mathbf{r}^{tot}$ is the total structural variation due the M mutations, $d\mathbf{r}^{mutm}_{wt}$ is the structural variation produced by the mutation $m$ (considering $\mathbf{K}^{wt}$), and $\mathbf{fmutm}$ is the vector of force of the $m$ mutation.

{\bf Model parameters: }
To completely specify the model, we must choose $\mathbf{K}_{wt}$ and $\mathbf{f}$. There are a number of alternative ENMs.Specifically, we used the ``Anisotropic Network Model'' (ANM). Following this model, we gave a spring force constant of 1 to sites at a distance $\leq$ 10 $\AA$ and a spring force constant of 0 to sites at a distance > 10 $\AA$. We selected 10 $\AA$ as the cut off value after optimization using a range of values form 8 to 12 (data not shown). This cut off value is also the most commonly used.
To calculate $\mathbf{f}$, given a mutation at a site $l$, each site $i$ in contact with $l$ is assigned a force directed along the $l-i$ contact and site $l$ is assigned a reaction force. To simulate random mutations, the magnitudes of $f_{li}$ were randomly picked from a uniform distribution in $[-f_{max},f_{max}]$. The forces for different contacts were picked independently. Since $f_{max}$ does not affect the results, we set
$f_{max} = 2$. We can think of the range $[-f_{max}, f_{max}]$ as a continuous approximation of the perturbations introduced by the mutations, covering
from mutations between physicochemically similar amino acids ($f \approx 0$) up to mutations between very different amino acids ($f \leq f_{max}$).

@@

{\it Selection of the reference protein:}
For each family of proteins we selected a reference ``wilt type'' protein, which we consider is the most structurally representative protein of the family. To get this protein, we first calculated the average structure of each multiple alignment. Then, we calculated the ``Mean Square Deviation'' (MSD) between the structure of each protein of the family and the obtained average structure. Finally, we selected as the reference protein the one with the lower value of MSD.

{\it Alignments analysis:}
For each family of proteins, the reference protein was aligned with each of the other proteins of the family. Then, the aligned and nonaligned sites, their coordinates, and the sequence identity for each pair of proteins were obtained.

{\bf Theoretical datasets:}
In order to study the effect of natural selection on structural divergence, for each family of proteins, we generated two datasets of simulated mutants. In one dataset, we applied natural selection and, on the other dataset, we did not consider natural selection:

{\it Mutants with natural selection (NS = T):} we applied natural selection at the sequence level. To determine which sites to mutate accounting natural selection, we followed the stress model, by which probabilities of mutation are assigned for each site of the reference protein:

\begin{equation}
p^{mut}_{i} = 1 - (beta \times CN_{i})
\end{equation}

Were $CN_{i}$ is the number of contacts (sites with a distance $\leq$ 10 $\AA$ in this case) of site $i$. 
To select the number of sites to mutate, we took information from the alignments. For each pair comprising the reference protein and one of the other proteins of the family, we obtained the average sequence identity (\% ID) within the aligned region. Then, the number of mutated sites for this pair is:

\begin{equation}
nsites^{mut} = (100 - \% ID) \times naa / 100
\end{equation}

Being $naa$ the number of aminoacids of the reference protein. 
Then, for each of these pairs, we generated 10 mutants of $nsites^{mut}$, where the mutated sites are the ones with more probabilities to mutate following the stress model.

In this way, if the family has $P$ proteins (plus the reference protein), we obtained a simulated dataset of $10 \times P$  members.

{\it Mutants without natural selection (NS = F):} For each pair comprising the reference protein and one of the other proteins of the family, we obtained the number of sites to mutate as mentioned before and we mutated randomly this number of sites. For each of these pairs, we generated 10 mutants. Hence, again, if the family has $P$ proteins plus the reference protein, we obtained a simulated dataset of  $10 \times P$  members.

\section{Results and discussion}

\section{Conclusion}
